% ICML submission — compile with pdflatex.
% Swap in the official icml2026.sty style file for submission; this preamble
% approximates the venue format with the article class.
\documentclass[10pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{booktabs,graphicx,microtype}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue]{hyperref}
\usepackage{times}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\linespread{1.02}
\title{Phase Transitions in Price-Weighted Routing Games: Learning Dynamics at the Knife Edge}
\author{Tarun Chitra\\Gauntlet \texttt{tarun@gauntlet.xyz}}
\date{}
\begin{document}
\maketitle

CI-enforced theory tests in the public `orcap` repository.*

\begin{abstract}
AI inference marketplaces route trillions of tokens per month through
a softmax over provider prices with a \emph{published} exponent
(selection probability $\propto p^{-a}$; deployed $a=2$). We study
independent Q-learning across this one-parameter family of pricing
games. The family has an exact phase structure:
symmetric equilibrium price $p^*=c\,a(n-1)/(a(n-1)-n)$ diverges on the
critical line $a(n-1)=n$, an entry-proof markup floor 1/a holds
everywhere, and the deployed parameter places two-provider markets
exactly at criticality. Our object of study is the *interaction of
learning dynamics with this phase structure*. Empirically (all runs
seeded, manifested, and gated behind a pre-registered validation of the
environment against a live market panel --- including an untargeted moment,
demand elasticity --0.65 $\pm$ 0.35 simulated vs --0.78 measured): (1) learned
prices are a smooth, strictly decreasing function of a that tracks the
equilibrium correspondence away from criticality but \textbf{regularizes the
singularity} --- near the critical line the profit gradient toward higher
prices vanishes at rate governed by h(p) $\to$ 1 and $\varepsilon$-greedy Q cannot climb
it, so learners undershoot divergent equilibria and overshoot interior
Nash ($\Delta$ = 0.31 at a = 2.5); (2) a measured history-dependent steering
penalty (weight $\times$0.17 on recent price cutters, ~7-epoch memory)
restructures the learning problem --- we prove it deters cutting for any
discount below $\delta^\dagger$ = 0.9895 and observe it flip learners from
bottom-of-book pricing to the menu ceiling in 4/4 calibrated markets
(+81\% on the flow-dominant quote where the measured conditional binds);
(3) learners systematically fail to discover the high-frequency
micro-adjustment strategy present in the real market (a 4.9$\times$ share tax
under the deployed steering makes it dominated), and converge to exact
price ties at the market's focal anchor without any salience assumption ---
the real market's two most distinctive regularities are attractors of the
learning dynamics. We further evaluate candidate mechanisms with learning
agents: a thickness-adaptive exponent removing criticality, and a
verified-quality weighting whose closed-form threshold $b^*=0.63$
separates learned quality-shading from learned quality provision. Our results suggest that deployed routing rules are a natural
laboratory where learning theory, mechanism design, and live-market
validation meet---and that near singular mechanism parameters, learned
play is predictable precisely where equilibrium analysis is not.

\end{abstract}

\section{Introduction}

Learning in games meets its cleanest applied test where a mechanism is
(i) deployed at scale, (ii) documented exactly, and (iii) instrumented.
AI inference routing is such a mechanism: a softmax over provider prices
with a published exponent decides, request by request, who serves each
call to a shared open-weight model. The exponent is an inverse
temperature; the induced game is logit Bertrand --- the very testbed of the
algorithmic-collusion literature (Calvano et al., 2020) --- except that
here the demand system is not a modeling choice: it is the product
documentation.

This paper treats the routing exponent as the control parameter of a
game family and asks how independent learners behave across it,
especially at its singular points. The family has a critical line
$a(n-1)=n$ where the symmetric equilibrium correspondence diverges to
the menu ceiling; the deployed parameter sits on it for duopolies. Away
from statistical mechanics analogies, this is a concrete learning
question: what does $\varepsilon$-greedy tabular Q do when the stage game's best-
response dynamics point toward an equilibrium at infinity along a
vanishing gradient?

\textbf{Contributions.} (1) A complete characterization of the game family's
phase structure with CI-verified closed forms, including an entry-proof
elasticity bound (share elasticity $\le$ a) and a measured-elasticity finite
version of the critical divergence (duopoly $p^*=41c$ at the marketplace's
measured outside elasticity). (2) A calibrated environment with a
pre-registered validation gate; behavioral species fitted from a live
panel; counterfactuals code-locked behind the gate. (3) The learning
phenomenology of the phase structure: smooth regularization of the
singularity; overshoot (supra-Nash, $\Delta$ up to 0.47) in the disciplined
phase; undershoot at criticality; mechanistic explanation via the
vanishing gradient of h. (4) Steering as reward shaping: a measured
penalty parameter, a deterrence theorem with a patience boundary, and
its empirical effect on learned play in calibrated markets. (5)
Mechanism evaluation with learners: adaptive temperature; quality
weighting with a closed-form learnability threshold.

 The router's softmax makes "inverse temperature" literal, not
metaphorical: allocation is a Gibbs measure with energy $a\log p$. The
phase-transition language below is the exact behavior of the equilibrium
correspondence, not an analogy.

\section{The game family}

n providers; prices $p_i\in(0,\bar p]$; router selects i with probability
$s_i=p_i^{-a}/\sum_j p_j^{-a}$; profit $\pi_i=s_i(p_i-c_i)$. Lemma 1:
own-price share elasticity $-a(1-s_i)\in(-a,0)$. Lemma 2 (single
crossing): best responses are unique --- root of $h(p)=a(1-s)(p-c)/p=1$
or the ceiling. Theorem 1: interior symmetric equilibrium
$p^*=c\,a(n-1)/(a(n-1)-n)$ iff $a(n-1)>n$, ceiling otherwise; Lerner floor
1/a in any interior equilibrium (any n, any outside option); with
measured outside elasticity $\varepsilon$: $p/(p-c)=a(n-1)/n+|\varepsilon|/n$ (duopoly at
deployed parameters: 41c). Theorem 2 (steering): penalty $\times\theta$ for $M$ epochs
after a cut $\Rightarrow$ optimal deviation $c+\sqrt{c^2+\theta/W}$; deterrence iff $\theta\le\theta^*$
($\in$ [0.81, 1] at calibrated configurations, measured $\theta$ = 0.17); patience
boundary $\delta^\dagger$ = 0.9895; perpetual-cutter share tax $\theta/(\theta+n-1)$ vs $1/n$.
Proofs in Appendix A; 13 closed forms CI-tested.

\section{Environment and validation}

(As in the companion papers; summarized.) Provider agents are behavioral
species fitted on the live panel's train window (adopters / static
undercutters / active undercutters / premium; split-sample validated);
demand replays the panel's AR(1)-diurnal process; costs are identified
bands from a capital-structure registry and the GPU spot book. The
pre-registered gate requires ten moments to match (distance $\le$ 0.04,
achieved 0.019, 20 seeds, deterministic replication), with the evidential
weight on two emergent moments: untargeted flow elasticity (--0.65 $\pm$ 0.35
vs --0.78) and out-of-sample adopter-atom persistence (0.83 vs 0.834).
Learning experiments use expected-allocation rewards (exact under
non-binding capacity; the request-level kernel supplies sampling noise in
the validation runs) and Calvano hyperparameters ($\alpha$ = 0.15, $\gamma$ = 0.95,
exponentially decaying $\varepsilon$), with convergence defined by greedy-policy
stability.

\section{Learning across the phase diagram}

\paragraph{4.1 The dial.} Converged mean prices (n = 3, 8 seeds/point):
uniform 1.53; a = 1: 1.30; a = 2: 0.94; a = 4: 0.60; a $\ge$ 8: grid floor.
Learned price is smooth and strictly decreasing in a --- the mechanism's
comparative statics survive bounded rationality everywhere.

\paragraph{4.2 At criticality.} For n = 3 the critical point is a = 1.5. There,
theory says the symmetric equilibrium is the ceiling (1.6); learners
reach 1.098 $\pm$ small. Mechanism: near criticality $d\pi/dp\propto 1-h(p)$ with
h $\to$ 1, so the drift toward the ceiling is arbitrarily weak relative to
exploration noise; tabular Q with decaying $\varepsilon$ performs a random walk on a
nearly flat profit landscape and settles far below the divergent
equilibrium. Conversely at a = 2.5 (interior Nash 0.5) learners hold
0.877 --- supra-Nash by $\Delta$ = 0.31: in the disciplined phase the landscape
has curvature and the classic Calvano reward-punishment channel operates.
\textbf{Learning friction regularizes the equilibrium correspondence:
undershoot where it diverges, overshoot where it is interior.} We view
this as the paper's most transferable observation: mechanism analysis
that stops at the equilibrium correspondence mispredicts learned play in
\emph{both directions}, in a sign pattern that a vanishing-gradient argument
predicts.

\paragraph{4.3 What is not learned.} Substituting a learner into each behavioral
slot (8/8 seeds unanimous): micro-adjustment (>1 reprice/day) never
emerges --- under the deployed steering it carries a 4.9$\times$ share tax and is
dominated (Theorem 2(iv)); the learner instead parks at a rigid
below-anchor price. With no static undercutter present, the learner
converges to the anchor price \emph{exactly}: endogenous tie formation. The
real market's tie atom (45\% of same-model pairs, 3.4$\times$ its mechanical
null) and its bifurcation into rigid-vs-micro-adjusting technologies are
thus both attractors of learning under the deployed mechanism --- an
out-of-sample style of validation we find more convincing than moment
matching: the environment reproduces *qualitative regularities it was
never told about.*

\paragraph{4.4 Steering as reward shaping.} The measured penalty is a
history-dependent, asymmetric transformation of the reward: cuts move the
agent into a taxed state for M epochs. Effect on learned play: stylized
world --- learner jumps 0.72 $\to$ ceiling (8/8 seeds), market +18\%.
Calibrated markets --- ceiling 4/4 under the broad rule; under the
strictly-measured cheapest-only conditional, the effect concentrates at
the book's bottom: +81\% on the flow-dominant quote where the learner is
cheapest, escape otherwise (both variants reported; they bracket the
unobserved rule). The penalty converts "undercut" from the greedy action
into a trap state; a $\gamma$ = 0.95 learner ($\delta^M$ $\approx$ 0.70 $\ll$ $\delta^\dagger$ = 0.9895) can
never profit from entering it.

\section{Mechanism evaluation with learning agents}

\paragraph{5.1 Objective frontier under cost heterogeneity.} Two calibrated cost
types (owned-capacity 0.10 $\times$2; spot-dependent 0.50 $\times$3). Theory FOCs and
learner runs [E-MECH1] agree on the ordering: welfare $\uparrow$ in a (better
cost-sorting), ad-valorem platform revenue $\downarrow$ in a, spot-type viability
$\downarrow$ 0 (the resilience cost of sharp routing). The deployed (a = 2,
ad-valorem fee) pair is revenue-serving, not welfare-serving; a
thickness-adaptive a\emph{(n) = n/($\ell$}(n--1)) pins markups and removes the
critical line. \textbf{5.2 Quality learnability threshold.} Binary quality
action; shading saves $\Delta$ = 0.08 at hidden damage d = 0.2. Weight
$w=q^b p^{-a}$: all-high-quality is an equilibrium iff $b\ge b^*=0.63$ (closed
form). E-MECH2 (5 seeds, 600k epochs): learned hi-quality share rises 0.27 (b = 0, the deployed quality-blind rule) $\to$ 0.40 (b = 0.5) $\to$ 0.67 (b = 1) $\to$ 0.53 (b = 2). The direction matches the theory --- quality weighting shifts learned play toward quality provision, with the largest jump crossing $b^*$ --- but the bifurcation is noisy rather than sharp: the quality margin ($\Delta$ = 0.08) is small relative to price-learning noise in the doubled action space, and at b = 2 quality-weight and price competition interact. We report this as directional confirmation of the closed-form threshold, not a sharp learnability boundary. The q signal exists operationally (our
deployed daily evaluation probes: graded benchmarks and greedy-output
hashes per provider).

\section{Related work}

Algorithmic pricing and collusion: Calvano et al.; Klein (sequential
moves); Asker et al. (learning protocols); Abada--Lambin (exploration-
driven pseudo-collusion); Hansen--Misra--Pai (misspecified bandits); Deng--
Schiffer--Bichler (deep RL nearer Nash). Our departure: the demand system
is deployed and documented, the environment is validation-gated against a
live market, and the phase-structure/learning interaction is the object.
Steering: Johnson--Rhodes--Wildenbeest (anti-collusive steering design) ---
we analyze the measured inverse. Mechanism/learning interfaces: reward
shaping, potential-based equivalences; the cut-penalty is a naturally
occurring non-potential shaping with measured parameters. Marketplace
empirics: Demirer et al. (OpenRouter). Softmax/temperature in games:
logit QRE (McKelvey--Palfrey) --- our learners' smooth dial echoes QRE
comparative statics, but the regularization sign pattern at criticality
is, to our knowledge, undocumented.

\section{Limitations}

Tabular Q with one hyperparameter suite (Calvano's), chosen for
comparability; a systematic hyperparameter/algorithm sweep (optimism,
UCB, PPO) is the natural next step and prior evidence suggests deep RL
lands nearer Nash --- which would sharpen, not reverse, \S4.2's contrast.
Expected-allocation rewards; symmetric-equilibrium theory; steering $\theta$
measured for one buyer tier; short calibration panel with a registered
30-day re-estimation that automatically reopens claims on sign flips.

\section{Reproducibility statement}

Public repository with: frozen calibration bundles (train/holdout
recorded), pre-registration documents with dated addenda, per-run
manifests (commit, bundle hash, seed tree, source fingerprints),
46 tests including 13 enforcing every closed form in \S2 against continuum
numerics, and the live-capture workflows that generated the panel.

\section{Appendix A: proofs.  Appendix B: run tables (E-SIM1--4b, E-MECH1--2).}

\end{document}
